A thesis draft

### 2018-11-21 res/sections/81-k8s.tex
English fix

\chapter{Kubernetes}\label{kubernetes}

In this appendix we provide a basic guide to Kubernetes concept and
installation. This is not thought as an complete overview of functionalities but
there are only reported a description of concept an feature used in the
implementation of our solution.

\section{Kubernetes basic concepts}\label{kubernetes-basic-concepts}

\subsection{Pod}\label{pod}

A pod is the most basic building block for Kubernetes. A pod usually
contains one or more container (if the containers are highly coupled),
even if Kubernetes it's able to manage different virtualization technologies
(e.g.
Virtual Machines\footnote{\url{https://github.com/kubevirt/kubevirt}}).
Pods have a life cycle, where it can assume these
states\footnote{
\url{https://kubernetes.io/docs/concepts/workloads/pods/pod-lifecycle/}}:
% TODO davide must delete from test
\begin{itemize}
\item
  \textbf{Pending}: The Pod has been accepted by the Kubernetes system,
  but one or more of the Container images has not been created. This
  includes time before being scheduled as well as time spent downloading
  images over the network, which could take a while.
\item
  \textbf{Running}: The Pod has been bound to a node, and all of the
  Containers have been created. At least one Container is still running,
  or is in the process of starting or restarting.
\item
  \textbf{Succeeded}: All Containers in the Pod have terminated in
  success, and will not be restarted.
\item
  \textbf{Failed}: All Containers in the Pod have terminated, and at
  least one Container has terminated in failure. That is, the Container
  either exited with non-zero status or was terminated by the system.
\item
  \textbf{Unknown}: For some reason the state of the Pod could not be
  obtained, typically due to an error in communicating with the host of
  the Pod.
\end{itemize}

A pod can have different conditions too:
% TODO davide must delete from test
\begin{itemize}
\item \textbf{PodScheduled}: the Pod has been scheduled to a node;
\item \textbf{Ready}: the Pod is able to serve requests and should be added
  to the load balancing pools of all matching Services;
\item \textbf{Initialized}: all init containers\footnote{
  \url{https://kubernetes.io/docs/concepts/workloads/pods/init-containers}}
  have started successfully;
\item \textbf{Unschedulable}: the scheduler cannot schedule the Pod right
  now, for example due to lacking of resources or other constraints;
\end{itemize}

Each pod can have different QoS classes:
\begin{description}
  \item [Guaranteed:] every container in the Pod must have a memory and CPU
  limits and memory and CPU requests, and they must be the same;
  \item [Burstable:] the Pod does not meet the criteria for QoS class
  Guaranteed and at least one container in the Pod has a memory or CPU request;
  \item [BestEffor:] the containers in the Pod must not have any memory or CPU
  limits or requests.
\end{description}

\subsection{Service}\label{service}

Since pods for their nature are highly volatile, applications that use
them can not rely on their IP address. To solve this problem, the
Service implementation in Kubernetes assure a stable IP, even if
different pods in this service crash and gets restarted by a ReplicaSet
(witch will not be explain in this guide, but it basically scale up and
down pods\footnote{
\url{https://kubernetes.io/docs/concepts/workloads/controllers/replicaset/}})

@@

\subsection{Volume}\label{volume}

When Pods saves data on a node, this information is designated to perish.
Kubernetes offers a way to save all data and to possibly make it
available to all the pods replicas, offering a transparent layer where
pods are not aware of working with information shared other pods, even located
in other nodes. This functionality is called Volume, and there are many
different drivers, that range from a simple local storage to glusterfs.

\subsection{Namespace}\label{namespace}

@@

A node is the building block for a Kubernetes cluster. Everyone must have
at least the kubelet daemon (to manage Kubernetes functionalities) and a
container manager (e.g. Docker Engine, VirtualBox) running. A node runs pods
and can communicate with other nodes in the same cluster

@@

\subsection{Master}\label{master}

The \emph{master node} is the one that starts a cluster, and that manage the
slaves. Only some pods can run n the master node (e.g. the dashboard).

@@

\subsection{Slave}\label{slave}

A \emph{slave} is a node that run the pods that the master schedule. It can
also be part of a \emph{PersistantVolume} storage (like GlusterFS) and if it
goes down, another slave will take care of the pods that where running on
it. Slaves can join existing cluster via a given token.

\subsection{Networking}\label{networking}

@@

Every nodes in the cluster runs a pod called \texttt{kubeproxy}, that
communicates with the \texttt{kube-api} pod (running on master) and
the \texttt{kube-dns} (for Kubernetes \textless{}=1.9.x) or
\texttt{CoreDNS} (for Kubernetes \textgreater{}=1.10.x). \texttt{kubeproxy}
assure transparent communication between pods in the same cluster, and
\texttt{kube-api} allows to call Kubernetes services. kube-dns/CoreDNS manage
DNS request coming from the pods, differentiating the one that goes outside
the cluster from the other that point internally\footnote{Additional
information about Flannel, Kubernetes and Docker networking
can at:\\
\sloppy
\url{https://www.sdxcentral.com/cloud/containers/definitions/what-is-coreos-flannel-definition/}}.

\begin{figure}[htbp]
\centering
\includegraphics[scale=0.3]{flannelredirect}
\caption{And example of how flannel redirect packages between pods}
\end{figure}

@@

Managing compute and storage are totally different problems and in order
to abstract the last one Kubernetes relies on \textbf{PersistantVolume}
and \textbf{PersistantVolumeClaim} resources.

A PersistantVolume (\textbf{PV}) is something similar to a node, in fact
is a piece of storage provided by the administrator: it captures the
details of the implementation of the storage and has a lifecycle that is
independent of any individual pod. To not expose developers to storage
boring details admins can expose different PV types through
\textbf{StorageClasses}.

A PersistantVolumeClaim (\textbf{PVC}) instead is a request for storage
and it is similar a pod: instead of requesting memory and CPU, a PVC
requests certain size and access mode to storage.

@@

It is important to understand the difference on \textbf{Volumes} and
\textbf{PersistantVolumes}/\textbf{PersistantVolumeClaims}. The former is
something like Docker volumes, used for persistence, host-based or on external
block storage devices but a user does not expect resources be reserved before
using volumes. In persistent volumes and claims, there is a strict enforcement
of resource utilization dictated by the policy defined during the
creation of resources. Kubernetes pod can use claims as volumes.

\textbf{StorageClasses}\footnote{more information on different provisioners,
policies and parameters can be found here: \\\url{
https://kubernetes.io/docs/concepts/storage/storage-classes/
}} are a way to describe different type of storage
that can be used. Different classes can be used to describe different
levels of service or policies. PVCs that don't specify a particular class
can use the default one: to define it you can use annotation
\verb!"storageclass.kubernetes.io/is-default-class":"true"!, as for
instance:

@@

\begin{enumerate}
  \item \textbf{Provisioning}: there are 2 ways: statically or dynamically.
  \emph{Static} means that an administrator creates some PVs that are
  available for consumption. \emph{Dynamic} means that the cluster
  dynamically try to provide a volume that matches a PVC. This is done
  with \emph{StorageClasses}: a claim requests storage for a certain
  storage class that can be provisioned. The class \texttt{""} disable
  dynamic provisioning;
\item
  \textbf{Binding}: a user requests a certain amount of storage with a
  claim and a control loop in the master look for a PV that can bind the
  claim. In case of dynamic provisioning the loop always bind a PVC to a
  PV, otherwise user will always get at least what they asked for, but
  the volume may be larger. Claims can remain unbound if a matching
  volume does not exists.
\item
  \textbf{Using}: after the binding phase the pod start using the volume
  (PVC is described in the pod yaml
  \footnote{\url{https://kubernetes.io/docs/concepts/storage/persistent-volumes/\#claims-as-volumes}}
\item
  \textbf{Releasing}: when an application is done with using the volume,
  developers can delete the PVC objects through the API, releasing the
  claim. This step will initiate the reclamation process.
\item
  \textbf{Reclaiming}: this phase is defined as a policy by the
  administrators. The reclaim policy for a PersistantVolume tells the
  cluster what to do with the volume after it has been released of its
  claim. Persistent volumes can either be \emph{retained},
  \emph{recycled} or \emph{deleted}.
\end{enumerate}

@@

We performed our Kubernetes installation in a Openstack environment.
Since we used the Openstack of our school department, the administrator
have given us a virtual machine, Openstack-CLI, to perform
operations. From now, our scripts assume you're in a machine with access
to Openstack, and that you have sourced your openrc.sh script before
continuing.

@@

First all the software needed must to be installed an for this the
script \\\verb!install_kubectl_repo.sh!
\footnote{\url{https://raw.githubusercontent.com/Augugrumi/init-script/centos/kubernetes/install_kubectl_repo.sh}}
was created. This only call another script and wait until all machine
are ready. The other script is \verb!bootstrap.sh!
\footnote{\url{https://raw.githubusercontent.com/Augugrumi/vagrantfiles/oldversion/kubernetes/centos/bootstrap.sh}}
whose aim is to configure the machine and the install the necessary
software.\\
Boostrap script first \textbf{disable SELinux} on all machines (needed
for Kubernetes networking)

\begin{lstlisting}
sudo setenforce 0
sudo sed -i --follow-symlinks 's/SELINUX=enforcing/SELINUX=disabled/g' /etc/sysconfig/selinux
\end{lstlisting}

then \textbf{enable br\_netfilter} module (needed for Kubernetes
networking)

\begin{lstlisting}
sudo modprobe br_netfilter
sudo sh -c 'echo "1" > /proc/sys/net/bridge/bridge-nf-call-iptables'
\end{lstlisting}

\textbf{Switch off the swap} by the commands

\begin{lstlisting}
sudo swapoff -a
sudo sed -i '/swap/d' /etc/fstab
\end{lstlisting}

And finally add \texttt{rbd}, \texttt{ip\_vs}, \texttt{ip\_vs\_rr},
\texttt{ip\_vs\_wrr}, \texttt{ip\_vs\_sh}, \texttt{dm\_snapshot},
\texttt{dm\_mirror} and \texttt{dm\_thin\_pool} to modules that will be
loaded to machine startup.

\begin{lstlisting}
sudo sh -c 'echo -e "rbd\nip_vs\nip_vs_rr\nip_vs_wrr\nip_vs_sh" >  /etc/modules-load.d/ip_vs.conf'
sudo sh -c 'echo -e "dm_snapshot\ndm_mirror\ndm_thin_pool" >  /etc/modules-load.d/gluster.conf'
\end{lstlisting}

Note: Last 3 modules will be useful in the case that you want to install
GlusterFS on your machines.

In the end the script install \texttt{docker},
\texttt{device-mapper-persistent-data}, \texttt{lvm2}, \texttt{kubelet},
\texttt{kubectl} and \texttt{kubeadm} and start (and enable)
\texttt{docker} and \texttt{kubelet} services.

After rebooting all the machine, the installation go on adding in the
\texttt{/etc/hosts} file of all node that will be part of the Kubernetes
cluster the IP and the name of all the nodes.

After the installation continues in the \textbf{master} node, with the script
\verb!on_master.sh!
\footnote{\url{https://raw.githubusercontent.com/Augugrumi/init-script/centos/kubernetes/on_master.sh}}.
We run \texttt{kubeadm} by the command

\begin{lstlisting}
sudo kubeadm init --apiserver-advertise-address=<ip-of-the-master> --pod-network-cidr=10.244.0.0/16 --ignore-preflight-errors cri | grep "kubeadm join" > /home/centos/joincommand
\end{lstlisting}

Saving part of the output that will be necessary for other nodes to join
the cluster.\\
Note 1:\texttt{-\/-pod-network-cidr} depends on which agent you want to
use to configure Kubernetes network. We choose Flannel so we put
\texttt{10.244.0.0/16}.\\
Note 2: we don't know at the moment why some preflight errors was
launched by the command and in order to do so (since everything seems
working) we added \texttt{-\/-ignore-preflight-errors\ cri} to go on.\\
This save in the \texttt{joincommand} file the command to be run on
slave node to join the cluster.

And after give the following commands to create the folder needed by
Kubernetes

\begin{lstlisting}
mkdir -p $HOME/.kube
sudo cp -i /etc/kubernetes/admin.conf $HOME/.kube/config
sudo chown $(id -u):$(id -g) $HOME/.kube/config
\end{lstlisting}

Then wait for all pods that are ready
(\texttt{kubectl\ get\ pods\ -\/-all-namespaces}).

Following, install Flannel\footnote{\url{https://github.com/coreos/flannel}}
for the networking and wait that it is up:

\begin{lstlisting}
kubectl apply -f https://raw.githubusercontent.com/coreos/flannel/master/Documentation/kube-flannel.yml
kubectl apply -f https://raw.githubusercontent.com/coreos/flannel/master/Documentation/k8s-manifests/kube-flannel-rbac.yml
\end{lstlisting}

And then it is possible to setup the dashboard to access Kubernetes resources:

\begin{lstlisting}
kubectl create serviceaccount dashboard -n default
  kubectl create clusterrolebinding dashboard-admin -n default \
   --clusterrole=cluster-admin \
   --serviceaccount=default:dashboard
  kubectl get secret $(kubectl get serviceaccount dashboard -o jsonpath="{.secrets[0].name}") -o jsonpath="{.data.token}" | base64 --decode >  secret.txt
  screen -dmS kubectl_proxy_screen bash
  screen -S kubectl_proxy_screen -X stuff "kubectl proxy
  "
\end{lstlisting}

With this command we make the dashboard accessible on port 8001 of the
master node at this URL
\begin{verbatim}
https://<master-ip>:<apiserver-port>/api/v1/namespaces/kube-system/services/https:kubernetes-dashboard:/proxy/}
\end{verbatim}

The previous command also the token to access the dashboard on
\texttt{secret.txt} file on the home folder of the master node.

The last thing to do is to make other node join the cluster, using the
\texttt{joincommand} file that we created before.

\section{Persistent Storage}\label{persistent-storage}

\subsection{Prerequisites}\label{prerequisites}

@@

\subsection{GlusterFS}\label{glusterfs}

@@

To enable persistent storage in Kubernetes with Glusterfs we need at
least three nodes. As we've already written above, we deployed our
Kubernetes cluster using CentOS 7.\\
Note: this procedure was tested on \texttt{Kubernetes\ v1.11.1.}

Since we're running our cluster on a Openstack environment, we attached
one volume per VM. If you're in our shoes, just create some empty
volumes and attach them to nodes you want to configure with glusterfs.

@@

For all the nodes you installed glusterfs you need to apply a particular
label in you Kubernetes cluster. In a terminal, type:

\begin{lstlisting}
kubectl label node <node name> storagenode=glusterfs
\end{lstlisting}

According to the \emph{gluster-kubernetes setup guide}
\footnote{\url{https://github.com/gluster/gluster-kubernetes/blob/master/docs/setup-guide.md}},
you should also check that the following ports are reachable:

\begin{itemize}
\item 2222
\item 24007
\item 24008
\item 49152-49251
\end{itemize}

On top of that, you'll need to grant access to your root account while
installing all the stuff we need. In CentOS, if like us you're using the
CloudImg image, you need to access every machine and type:

\begin{lstlisting}
sudo su
cd /root/
nano .ssh/authorized_keys
# edit the first line in the follwing way: from the sub-string starting
# as "ssh-rsa" until the end put the content in a new line, while comment
# the whole line
nano /etc/ssh/sshd_config # un-comment "PermitRootLogin: yes"
systemctl restart sshd
\end{lstlisting}

At this point, you need to clone \texttt{kubernetes-glusterfs}, the repository
that contains a script that will figure out all the necessary stuff
(e.g. heketi), \texttt{gk-deploy}.

\begin{lstlisting}
git clone https://github.com/gluster/gluster-kubernetes.git
cd gluster-kubernetes/deploy
\end{lstlisting}

Now we need to define the topology. To do so, a \texttt{topology.json}
file must be created, from the sample \texttt{topology.json.sample}, and edit to
fit your topology. As in the following example:

\begin{lstlisting}
{
 "clusters": [
   {
     "nodes": [
       {
         "node": {
           "hostnames": {
             "manage": [
               "192.168.29.26"
             ],
             "storage": [
               "192.168.29.26"
             ]
           },
           "zone": 1
         },
         "devices": [
           "/dev/vdb"
         ]
       },
       {
         "node": {
           "hostnames": {
             "manage": [
               "192.168.29.19"
             ],
             "storage": [
               "192.168.29.19"
             ]
           },
           "zone": 1
         },
         "devices": [
           "/dev/vdb"
         ]
       },
       {
         "node": {
           "hostnames": {
             "manage": [
               "192.168.29.22"
             ],
             "storage": [
               "192.168.29.22"
             ]
           },
           "zone": 1
         },
         "devices": [
           "/dev/vdb"
         ]
       }
     ]
   }
 ]
}
\end{lstlisting}

Now it is possible to deploy a Gluster cluster. From the master the following
command must be run:

@@

At the end, if everything goes correctly, a \texttt{StorageClass} template have
to be created. To your Kubernetes through \texttt{kubectl}:

\begin{lstlisting}[language=bash]
kubectl create -f storageclass.yaml
\end{lstlisting}

Optionally, it can be set as default \texttt{StorageClass} with the following
command:

\begin{lstlisting}[language=bash]
kubectl patch -f storageclass.yaml \
-p'{"metadata": {"annotations":{"storageclass.kubernetes.io/is-default-class":"true"}}}'
\end{lstlisting}

\section{Ingress}\label{ingress}

Ingress makes service available outside the cluster. There are different ways
Kubernetes handle service reachability, and these are:

>>>>

\chapter{Kubernetes}
\label{kubernetes}

In this appendix we provide a basic summary to the Kubernetes concepts and
we will show a simple way to perform a basic installation. This is not thought 
as an complete overview of functionalities but there are only reported a 
description of concepts an features used in the implementation of our solution.

\section{Kubernetes basic concepts}
\label{kubernetes-basic-concepts}

\subsection{Pod}
\label{pod}

A Pod is the most basic building block for Kubernetes. A Pod usually
contains one or more container (if the containers are highly coupled),
even if Kubernetes is able to manage different virtualization technologies
(e.g. Virtual Machines\footnote{\url{https://github.com/kubevirt/kubevirt}}).
Pods have a life cycle, where it can assume these
states\footnote{
\url{https://kubernetes.io/docs/concepts/workloads/pods/pod-lifecycle/}}:
% TODO davide must delete from test
\begin{itemize}
\item \textbf{Pending}: The Pod has been accepted by the Kubernetes system,
  but one or more of the Container images have not been created yet. This
  includes time before being scheduled as well as time spent downloading
  images over the network, which could take a while.
\item \textbf{Running}: The Pod has been bound to a node, and all of the
  Containers have been created. At least one Container is still running,
  or is in the process of starting or restarting.
\item \textbf{Succeeded}: All Containers in the Pod have terminated with
  success, and they will not be restarted.
\item \textbf{Failed}: All Containers in the Pod have terminated, and at
  least one Container has terminated in failure. That is, the Container
  either exited with non-zero status or was terminated by the system.
\item \textbf{Unknown}: For some reason the state of the Pod could not be
  obtained, typically due to an error in communicating with the host of
  the Pod.
\end{itemize}

A Pod can have different conditions too:
% TODO davide must delete from test
\begin{itemize}
\item \textbf{PodScheduled}: the Pod has been scheduled to a node;
\item \textbf{Ready}: the Pod is able to serve requests and should be added
  to the load balancing pools of all matching Services;
\item \textbf{Initialized}: all init containers\footnote{
  \url{https://kubernetes.io/docs/concepts/workloads/pods/init-containers}}
  have started successfully;
\item \textbf{Unschedulable}: the scheduler cannot schedule the Pod right
  now, for example due to lacking of resources or other constraints;
\end{itemize}

Each Pod can have different QoS classes:
\begin{description}
  \item [Guaranteed:] every container in the Pod must have a memory and CPU
  limits and memory and CPU requests, and they must be the same;
  \item [Burstable:] the Pod does not meet the criteria for QoS class
  Guaranteed and at least one container in the Pod has a memory or CPU request;
  \item [BestEffor:] the containers in the Pod must not have any memory or CPU
  limits or requests.
\end{description}

\subsection{Service}
\label{service}

Since Pods for their nature are highly volatile, applications that use
them can not rely on their IP address. To solve this problem, the
Service implementation in Kubernetes assure a stable IP, even if
different Pods in this service crash and gets restarted by a ReplicaSet
(witch will not be explain in this guide, but it basically scale up and
down Pods\footnote{
\url{https://kubernetes.io/docs/concepts/workloads/controllers/replicaset/}})

@@

\subsection{Volume}
\label{volume}

When Pods saves data on a node, this information is designated to perish.
Kubernetes offers a way to save all data and to possibly make it
available to all the Pods replicas, offering a transparent layer where
Pods are not aware of working with information shared other Pods, even located
in other nodes. This functionality is called Volume, and there are many
different drivers, that range from a simple local storage to glusterfs.

\subsection{Namespace}
\label{namespace}

@@

A node is the building block for a Kubernetes cluster. Everyone must have
at least the kubelet daemon (to manage Kubernetes functionalities) and a
container manager (e.g. Docker Engine, VirtualBox) running. A node runs Pods
and can communicate with other nodes in the same cluster

@@

\subsection{Master}
\label{master}

The \emph{master node} is the one that starts a cluster, and that manage the
slaves. Only some Pods can run in the master node (e.g. the dashboard).

@@

\subsection{Slave}
\label{slave}

A \emph{slave} is a node that run the Pods that the master schedule. It can
also be part of a \emph{PersistantVolume} storage (like GlusterFS) and if it
goes down, another slave will take care of the Pods that where running on
it. Slaves can join existing cluster via a given token.

\subsection{Networking}
\label{networking}

@@

Every node in the cluster runs a Pod called \texttt{kubeproxy}, that
communicates with the \texttt{kube-api} Pod (running on master) and
the \texttt{kube-dns} (for Kubernetes \textless{}=1.9.x) or
\texttt{CoreDNS} (for Kubernetes \textgreater{}=1.10.x). \texttt{kubeproxy}
assure transparent communications between Pods in the same cluster, and
\texttt{kube-api} allows to call Kubernetes services. \texttt{kube-dns/CoreDNS} 
manage DNS requests coming from the Pods, differentiating the one that goes 
outside the cluster from the other that points internally\footnote{Additional
information about Flannel, Kubernetes and Docker networking can at:\\
\sloppy
\url{https://www.sdxcentral.com/cloud/containers/definitions/what-is-coreos-flannel-definition/}}.

\begin{figure}[htbp]
\centering
\includegraphics[scale=0.3]{flannelredirect}
\caption{And example of how flannel redirect packages between Pods}
\end{figure}

@@

Managing compute and storage are totally different problems and in order
to abstract the last one Kubernetes relies on the \textbf{PersistantVolume}
and the \textbf{PersistantVolumeClaim} resources.

A PersistantVolume (PV) is something similar to a node, in fact
is a piece of storage provided by the administrator: it captures the
details of the implementation of the storage and has a lifecycle that is
independent of any individual Pod. To not expose developers to storage
boring details admins can expose different PV types through
\textbf{StorageClasses}.

A PersistantVolumeClaim (PVC) instead is a request for storage and it is 
similar a Pod: instead of requesting memory and CPU, a PVC requests certain 
size and access mode to storage.

@@

It is important to understand the difference on \textbf{Volumes} and
\textbf{PersistantVolumes}/\textbf{PersistantVolumeClaims}. The former is
something like Docker volumes, used for persistence, host-based or on external
block storage devices but a user does not expect resources be reserved before
using volumes. In persistent volumes and claims, there is a strict enforcement
of resource utilization dictated by the policy defined during the
creation of resources. Kubernetes Pod can use claims as volumes.

\textbf{StorageClasses}\footnote{more information on different provisioners,
policies and parameters can be found here: \\\url{
https://kubernetes.io/docs/concepts/storage/storage-classes/
}} are a way to describe different type of storage
that can be used. Different classes can be used to describe different
levels of service or policies. PVCs that do not specify a particular class
can use the default one: to define it you can use annotation
\verb!"storageclass.kubernetes.io/is-default-class":"true"!, as for
instance:

@@

\begin{enumerate}
\item \textbf{Provisioning}: there are 2 ways: statically or dynamically.
  \emph{Static} means that an administrator creates some PVs that are
  available for consumption. \emph{Dynamic} means that the cluster
  dynamically try to provide a volume that matches a PVC. This is done
  with \emph{StorageClasses}: a claim requests storage for a certain
  storage class that can be provisioned. The class \texttt{""} disable
  dynamic provisioning;
\item \textbf{Binding}: a user requests a certain amount of storage with a
  claim and a control loop in the master look for a PV that can bind the
  claim. In case of dynamic provisioning the loop always bind a PVC to a
  PV, otherwise user will always get at least what they asked for, but
  the volume may be larger. Claims can remain unbound if a matching
  volume does not exists.
\item \textbf{Using}: after the binding phase the Pod starts using the volume
  (PVC is described in the Pod yaml
  \footnote{\url{https://kubernetes.io/docs/concepts/storage/persistent-volumes/\#claims-as-volumes}}
\item \textbf{Releasing}: when an application is done with using the volume,
  developers can delete the PVC objects through the API, releasing the
  claim. This step will initiate the reclamation process.
\item \textbf{Reclaiming}: this phase is defined as a policy by the
  administrators. The reclaim policy for a PersistantVolume tells the
  cluster what to do with the volume after it has been released of its
  claim. Persistent volumes can either be \emph{retained},
  \emph{recycled} or \emph{deleted}.
\end{enumerate}

@@

We performed our Kubernetes installation in a Openstack environment.
Since we used the Openstack of our school department, the administrators have 
given us a virtual machine, Openstack-CLI, to perform operations. From now, our 
scripts assume you're in a machine with access to Openstack, and that you have 
sourced your openrc.sh script before continuing.

@@

First of all the software needed must be installed and for this the script 
\\\verb!install_kubectl_repo.sh!
\footnote{\url{https://raw.githubusercontent.com/Augugrumi/init-script/centos/kubernetes/install_kubectl_repo.sh}}
was created. This only calls another script and waits until all the machines
are ready. The other script is \verb!bootstrap.sh!
\footnote{\url{https://raw.githubusercontent.com/Augugrumi/vagrantfiles/oldversion/kubernetes/centos/bootstrap.sh}}
whose aim is to configure the machine and to install all the necessary
software. The boostrap script first \textbf{disables SELinux} on all machines 
(needed for Kubernetes networking):
\begin{lstlisting}
sudo setenforce 0
sudo sed -i --follow-symlinks 's/SELINUX=enforcing/SELINUX=disabled/g' /etc/sysconfig/selinux
\end{lstlisting}

and then it enables the \verb!br_netfilter! module (needed for Kubernetes
networking):
\begin{lstlisting}
sudo modprobe br_netfilter
sudo sh -c 'echo "1" > /proc/sys/net/bridge/bridge-nf-call-iptables'
\end{lstlisting}

The swap is then switched off with the following commands:
\begin{lstlisting}
sudo swapoff -a
sudo sed -i '/swap/d' /etc/fstab
\end{lstlisting}

And it finally adds \texttt{rbd}, \texttt{ip\_vs}, \texttt{ip\_vs\_rr},
\texttt{ip\_vs\_wrr}, \texttt{ip\_vs\_sh}, \texttt{dm\_snapshot},
\texttt{dm\_mirror} and \texttt{dm\_thin\_pool} to the modules that will be
loaded to the machine startup.
\begin{lstlisting}
sudo sh -c 'echo -e "rbd ip_vs ip_vs_rr ip_vs_wrr ip_vs_sh" >  
/etc/modules-load.d/ip_vs.conf'
sudo sh -c 'echo -e "dm_snapshot dm_mirror dm_thin_pool" >  
/etc/modules-load.d/gluster.conf'
\end{lstlisting}

Note: the last 3 modules will be useful in case GlusterFs has to be installed 
in the system.

In the end the script installs \texttt{docker}, 
\texttt{device-mapper-persistent-data}, \texttt{lvm2}, \texttt{kubelet},
\texttt{kubectl}, \texttt{kubeadm} and it starts (and enables)
\texttt{docker} and \texttt{kubelet} services.

After rebooting all the machines, the installation goes on adding in the
\texttt{/etc/hosts} file of all nodes IP that will be part of the Kubernetes
cluster and the names of all the nodes.

After that the installation continues in the \textbf{master} node, with the 
script \verb!on_master.sh!
\footnote{\url{https://raw.githubusercontent.com/Augugrumi/init-script/centos/kubernetes/on_master.sh}}.
The following command initiates the Kubernetes cluster:
\begin{lstlisting}
sudo kubeadm init --apiserver-advertise-address=<ip-of-the-master> 
--Pod-network-cidr=10.244.0.0/16 --ignore-preflight-errors cri | grep "kubeadm 
join" > /home/centos/joincommand
\end{lstlisting}

\noindent Saving part of the output that will be necessary for other nodes to 
join the cluster. There are some considerations to keep in mind though:
\begin{enumerate}
 \item \texttt{-\/-pod-network-cidr} depends on which agent you want to
use to configure Kubernetes network. We choose Flannel so we put
\texttt{10.244.0.0/16};
 \item we don't know at the moment why some preflight errors was
launched by the command and in order to do so (since everything seems
working) we added \texttt{-\/-ignore-preflight-errors\ cri} to go on.\\
This save in the \texttt{joincommand} file the command to be run on
slave node to join the cluster.
\end{enumerate}

And after its necessary to give the following commands to create the folder 
needed by Kubernetes to properly work:
\begin{lstlisting}
mkdir -p $HOME/.kube
sudo cp -i /etc/kubernetes/admin.conf $HOME/.kube/config
sudo chown $(id -u):$(id -g) $HOME/.kube/config
\end{lstlisting}

Then is necessary to wait for all Pods to be ready
(\texttt{kubectl\ get\ pods\ -\/-all-namespaces}).

The next step is to install 
Flannel\footnote{\url{https://github.com/coreos/flannel}}
in order to get the networking working. Some time is necessay until the 
networking is up:
\begin{lstlisting}
kubectl apply -f https://raw.githubusercontent.com/coreos/flannel/master/Documentation/kube-flannel.yml
kubectl apply -f https://raw.githubusercontent.com/coreos/flannel/master/Documentation/k8s-manifests/kube-flannel-rbac.yml
\end{lstlisting}

And then it is possible to setup the dashboard to access Kubernetes resources:
\begin{lstlisting}
kubectl create serviceaccount dashboard -n default
  kubectl create clusterrolebinding dashboard-admin -n default \
   --clusterrole=cluster-admin \
   --serviceaccount=default:dashboard
  kubectl get secret $(kubectl get serviceaccount dashboard -o jsonpath="{.secrets[0].name}") -o jsonpath="{.data.token}" | base64 --decode >  secret.txt
  screen -dmS kubectl_proxy_screen bash
  screen -S kubectl_proxy_screen -X stuff "kubectl proxy
  "
\end{lstlisting}

Finally, with this command we make the dashboard accessible on port 8001 of the 
master node at this URL
\begin{verbatim}
https://<master-ip>:<apiserver-port>/api/v1/namespaces/kube-system/services/https:kubernetes-dashboard:/proxy/}
\end{verbatim}

The previous command also necessitate of the token to access the dashboard, 
that is available in the file \texttt{secret.txt}, present on the home of the 
master node.

The last thing to do is to make other nodes to join the cluster, using the
\texttt{joincommand} file that we created before.

\section{Persistent Storage}
\label{persistent-storage}

\subsection{Prerequisites}
\label{prerequisites}

@@

\subsection{GlusterFS}
\label{glusterfs}

@@

To enable persistent storage in Kubernetes with Glusterfs we need at
least three nodes. As we have already written above, we deployed our
Kubernetes cluster using CentOS 7\footnote{This procedure was tested on 
\texttt{Kubernetes\ v1.11.1.}}.

Since we are running our cluster on a Openstack environment, we attached
one volume per VM. If you're in our shoes, just create some empty
volumes and attach them to the nodes you want to configure with Glusterfs.

@@

For all the nodes where you installed Glusterfs you need to apply a particular
label in you Kubernetes cluster. In a terminal, type:
\begin{lstlisting}
kubectl label node <node name> storagenode=glusterfs
\end{lstlisting}

According to the \emph{gluster-kubernetes setup guide}
\footnote{\url{https://github.com/gluster/gluster-kubernetes/blob/master/docs/setup-guide.md}},
you should also check that the following ports are reachable:
\begin{itemize}
\item 2222
\item 24007
\item 24008
\item 49152-49251
\end{itemize}

On top of that, you will need to grant access to your root account while
installing all the stuff we need. In CentOS, if like us you are using the
\texttt{CloudImg} image, you need to access every machine and type:
\begin{lstlisting}
sudo su
cd /root/
nano .ssh/authorized_keys
# edit the first line in the follwing way: from the sub-string starting
# as "ssh-rsa" until the end put the content in a new line, while comment
# the whole line
nano /etc/ssh/sshd_config # un-comment "PermitRootLogin: yes"
systemctl restart sshd
\end{lstlisting}

At this point, you need to clone \texttt{kubernetes-glusterfs}, the repository
that contains a script that will figure out all the necessary stuff
(e.g. heketi), \texttt{gk-deploy}.
\begin{lstlisting}
git clone https://github.com/gluster/gluster-kubernetes.git
cd gluster-kubernetes/deploy
\end{lstlisting}

Now we need to define the topology. To do so, a \texttt{topology.json}
file must be created, from the sample \texttt{topology.json.sample}, and edit to
fit your topology, like in the following example:
\begin{lstlisting}
{
 "clusters": [
   {
     "nodes": [
       {
         "node": {
           "hostnames": {
             "manage": [
               "192.168.29.26"
             ],
             "storage": [
               "192.168.29.26"
             ]
           },
           "zone": 1
         },
         "devices": [
           "/dev/vdb"
         ]
       },
       {
         "node": {
           "hostnames": {
             "manage": [
               "192.168.29.19"
             ],
             "storage": [
               "192.168.29.19"
             ]
           },
           "zone": 1
         },
         "devices": [
           "/dev/vdb"
         ]
       },
       {
         "node": {
           "hostnames": {
             "manage": [
               "192.168.29.22"
             ],
             "storage": [
               "192.168.29.22"
             ]
           },
           "zone": 1
         },
         "devices": [
           "/dev/vdb"
         ]
       }
     ]
   }
 ]
}
\end{lstlisting}

Now it is possible to deploy a Gluster cluster. From the master the following
commands must be run:

@@

At the end, if everything goes correctly, a \texttt{StorageClass} template needs
to be created. To your Kubernetes through \texttt{kubectl} type:
\begin{lstlisting}[language=bash]
kubectl create -f storageclass.yaml
\end{lstlisting}

Optionally, it can be set as default \texttt{StorageClass} with the following
command:
\begin{lstlisting}[language=bash]
kubectl patch -f storageclass.yaml \
-p'{"metadata": {"annotations":{"storageclass.kubernetes.io/is-default-class":"true"}}}'
\end{lstlisting}

\section{Ingress}
\label{ingress}

Ingress allows services to be available from outside the cluster. There are 
different ways Kubernetes can handle service reachability, and these are:
\begin{description}
  \item [ClusterIP:] Exposes the service on a cluster-internal IP. Choosing this
  value makes the service only reachable from within the cluster. This
  is the default ServiceType.
  \item [NodePort:]
  \footnote{\url{https://kubernetes.io/docs/concepts/services-networking/service/\#nodeport}}
  Exposes the service on each Node's IP at a static port (the NodePort).
  A ClusterIP service, to which the NodePort service will route, is
  automatically created. You'll be able to contact the NodePort service,
  from outside the cluster, by requesting
  \texttt{\textless{}NodeIP\textgreater{}:\textless{}NodePort\textgreater{}}.
  \item [LoadBalancer:]
  \footnote{\url{https://kubernetes.io/docs/concepts/services-networking/service/\#loadbalancer}}
  Exposes the service externally using a cloud provider's load balancer.
  NodePort and ClusterIP services, to which the external load balancer
  will route, are automatically created.
  \item [ExternalName:]
  \footnote{\url{https://kubernetes.io/docs/concepts/services-networking/service/\#externalname}}
  Maps the service to the contents of the ExternalName field
  (e.g.~foo.bar.example.com), by returning a CNAME record with its
  value. No proxying of any kind is set up. This requires version 1.7 or
  higher of kube-dns.
\end{description}
